---
title: "104_gam_test"
format: 
  html:
    toc: true
    code-fold: true  
editor: visual
---

-   Use model s(x,y) + s(year)\
-   Try using data for all Europe\
-   Use [soap-film smoother](https://fromthebottomoftheheap.net/2016/03/27/soap-film-smoothers/) for the spatial effect

```{r}

knitr::opts_chunk$set(results = 'hold') # collect the results from a chunk  

```

## 1. Packages

```{r, include=FALSE}

# Many of these packages cn be loaded at once using library(tidyverse). (I just like to be specific.)
library(mgcv)
library(dplyr)       # tidyverse 
library(ggplot2)     # tidyverse
library(viridis)     # for scale_color_viridis  
library(forcats)     # tidyverse
library(purrr)       # tidyverse
library(glue)        # tidyverse, easier alternative to paste0
library(tictoc)      # tic() and toc() for timing   
library(terra)       # spatial functions  

country_borders <- readRDS("Input_data/country_borders.rda")

```

```{r}

knitr::opts_chunk$set(results = 'hold') # collect the results from a chunk  

```

## 2. Data

```{r}

dat <- readRDS("Input_data/WISE_GWNO3_sel2.rda") %>%
  mutate(log_no3 = log10(no3))

# Locations  
 <- dat %>% 
  count(countryCode, monitoringSiteName, x, y)

```

## 3. Test GAM with soap-film smoother

### Test 1: example from GAM help pages

-   Example from ?soap

```{r}


##########################
## simple test function...
##########################

fsb <- list(fs.boundary())
nmax <- 100

## create some internal knots...
knots <- data.frame(v=rep(seq(-.5,3,by=.5),4),
                    w=rep(c(-.6,-.3,.3,.6),rep(8,4)))

## Simulate some fitting data, inside boundary...
set.seed(0)
n <- 600
v <- runif(n)*5-1;w<-runif(n)*2-1
y <- fs.test(v,w,b=1)
names(fsb[[1]]) <- c("v","w")
ind <- inSide(fsb,x=v,y=w) ## remove outsiders
y <- y + rnorm(n)*.3 ## add noise
y <- y[ind];v <- v[ind]; w <- w[ind] 
n <- length(y)

par(mfrow=c(3,2))
## plot boundary only
plot(fsb[[1]]$v,fsb[[1]]$w,type="l")

## plot boundary with knot and data locations
plot(fsb[[1]]$v,fsb[[1]]$w,type="l")
points(knots, pch=20, col=2)
points(v,w,pch=".");

## Now fit the soap film smoother. 'k' is dimension of boundary smooth.
## boundary supplied in 'xt', and knots in 'knots'...
 
nmax <- 100 ## reduced from default for speed.
# str(v, 1)
# str(w, 1)
# str(y, 1)
# str(fsb, 2)
# str(knots, 1)
b <- gam(y ~ s(v, w, k=30, bs="so", 
               xt = list(bnd = fsb, nmax=nmax)), 
         knots=knots)

plot(b) ## default plot
plot(b,scheme=1)
plot(b,scheme=2)
plot(b,scheme=3)

vis.gam(b,plot.type="contour")


```

### Test 2: More realistic

-   Example from soap() but setting the knots using Gavin's method\
-   https://fromthebottomoftheheap.net/2016/03/27/soap-film-smoothers/

```{r}
#| results: hold

#
# Boundary
#
cat("------------ fs.boundary() ------------ \n")
fs.boundary() %>% str(1)

# Make fsb (list of lists)
fsb <- list(fs.boundary())

# Change names, in order to fit with data  
names(fsb[[1]]) <- c("v","w")
cat("\n------------ fsb ------------ \n")
fsb %>% str(2)

## plot boundary only
plot(fsb[[1]]$v, fsb[[1]]$w, type="l")
mtext("Boundary")

#
# Simulate some fitting data, inside boundary...
#

# Locations
set.seed(1)
n <- 600
v <- runif(n)*5-1; w <- runif(n)*2-1
plot(v, w)
mtext("Data locations")

# Data (without noise)
y <- fs.test(v, w, b=1)

# Plot data
dat_test1 <- data.frame(v=v, w=w, y=y)
ggplot(dat_test1, aes(v, w, color = y)) +
  geom_point() +
  scale_color_viridis(na.value = "red")  +
  ggtitle("Data, before removing outsiders")


# For removing outsiders  
ind <- inSide(fsb, x=v, y=w) 
str(ind)

## add noise
y <- y + rnorm(n)*.3 

# remove outsiders
y <- y[ind]; v <- v[ind]; w <- w[ind] 
n <- length(y)

#
# Plot data again
#
dat_test2 <- data.frame(v=v, w=w, y=y)
ggplot(dat_test2, aes(v, w, color = y)) +
  geom_point() +
  scale_color_viridis(na.value = "red")  +
  ggtitle("Data, after removing outsiders")

#
# Create knots following Gavin  
#
N <- 10
gx <- seq(min(v), max(v), len = N)
gy <- seq(min(w), max(w), len = N) + 0.1  # add 0.1 in order to have fewer points on the border
# set.seed(1)
# gx <- runif(N)*5-1; gy <- runif(N)*2-1
gp <- expand.grid(gx, gy)
names(gp) <- c("x","y")

# Check our "raw" knots
ggplot(gp, aes(x, y)) +
  geom_point() +
  ggtitle("Knots, raw")

# Use inSide() to weed out knots  

# This doesn't work - x and y must be vectors with the names used
#   in 'fsb' (!)
# sel <- inSide(fsb, x=gp$x, y=gp$y)
# This works - x and y must be vectors with the names used
v <- gp$x
w <- gp$y
sel <- inSide(fsb, x=v, y=w) 
plot(fsb[[1]]$v, fsb[[1]]$w, type = "l")
points(v, w, col = "grey50")
text(v[sel], w[sel], 1:sum(sel), col = "blue2")
mtext("Knots, before manual weeding")

# Make data frame for knots and start testing the GAM
# - manual weeding must take place, rinse and repeat until the GAM works  
knots <- data.frame(v = v[sel], w = w[sel])
# After error in the gam: "knot 1 is on or outside boundary", add:
knots <- knots[-1,]
# After error "knot 25 is on or outside boundary", add:
knots <- knots[-25,]
# Etc. - remove points one by one until
# Note that you also can get th error message:
#    Error in soap.basis(sd, x, y, film = TRUE, wiggly = TRUE, penalty = TRUE) : 
#    NA/NaN/Inf in foreign function call (arg 13)
# but it still is handled the same way, but find the knot number by examining the plot below
knots <- knots[-60,]
knots <- knots[-16,]
knots <- knots[-32,]
knots <- knots[-49,]
plot(fsb[[1]]$v, fsb[[1]]$w, type = "l")
points(knots$v, knots$w, col = "grey50")
text(knots$v, knots$w, 1:nrow(knots), col = "blue2")
mtext("Knots used in analysis")

# GAM - dimension of the PDE solution grid
nmax <- 100 

# GAM with soap smoother
cat("\n=====================  Input for the GAM ===================== \n")
cat("Note that the names 'v' and 'w' must be the same for 3 objects: dat_test2, fsb and knots \n")
cat("\n------------ dat_test2 ------------ \n")
str(dat_test2, 1)
cat("\n------------ fsb ------------ \n")
str(fsb, 2)
cat("\n------------ knots ------------ \n")
str(knots, 1)
mod <- gam(y ~ s(v, w, k=30, bs="so", 
               xt = list(bnd = fsb, nmax=nmax)),
         data = dat_test2,
         knots=knots)

plot(mod) ## default plot


```

### Test 3: prediction

```{r}

grid.v <- seq(min(fsb[[1]]$v), max(fsb[[1]]$v), length = 40)
grid.w <- seq(min(fsb[[1]]$w), max(fsb[[1]]$w), length = 40)
pdata <- expand.grid(v = grid.v, w = grid.w)
pdata2 <- transform(pdata[, 1:2], y = predict(mod, newdata = pdata))

# head(pdata2)

ggplot(pdata2, aes(v, w, color = y)) +
  geom_point() +
  scale_color_viridis(na.value = "red")  +
  ggtitle("Predicted y")

```

## 4. Map data for land

-   Europe, land, from Natural Earth website\
-   Cropping data, parly manually to

### Original data, cropped  
```{r}

# SpatVector from file
f <- "ne_50m_land/ne_50m_land.shp"
f <- "ne_110m_land/ne_110m_land.shp"
boundary_land_all <- vect(f)
# plot(boundary_land_all)

# crop with an extent
ex <- ext(-10, 50, 30, 73)
boundary_land <- crop(boundary_land_all, ex)
plot(boundary_land, main = "Original map (Mercator)")  


```

### Project to ETRS89  

```{r}

# help(package = "terra")

# Project to ETRS89  
crs_ETRS89_extended <- "+proj=lcc +lat_0=52 +lon_0=10 +lat_1=35 +lat_2=65 +x_0=4000000 +y_0=2800000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs +type=crs"
boundary_land2 <- project(boundary_land, crs_ETRS89_extended)
plot(boundary_land2, main = "Map in ETRS89 projection")  

```

### Create x,y coordinates in km  

```{r}

# Get coordinates  
boundary_coord <- crds(boundary_land2, df = TRUE)
names(boundary_coord) <- c("x_orig", "y_orig")

# Define x, y as km from lat = 27 & lon = -15

# Get the transformed coordinates for lat = 27 & lon = -15  
longitude <- c(-116.7, -120.4, -116.7, -113.5, -115.5, -120.8, -119.5, -113.7, -113.7, -110.7)
latitude <- c(45.3, 42.6, 38.9, 42.1, 35.7, 38.9, 36.2, 39, 41.6, 36.9)
lonlat <- cbind(-15, 27)
crs_longlat <- "+proj=longlat +datum=WGS84"
origo_point <- vect(lonlat, crs=crs_longlat)
origo_point_t <- project(origo_point, crs_ETRS89_extended)
origo_coord <- crds(origo_point_t)

# Define x, y 
boundary_coord$x <- (boundary_coord$x_orig - origo_coord[1,"x"])/1000
boundary_coord$y <- (boundary_coord$y_orig - origo_coord[1,"y"])/1000

```

### Make final map  

```{r}

# Test plots
if (FALSE){

  # IMPORTANT:
  # This plot is used for  selectiong points for boundary_coord2 and boundary_coord3
  plot(y~x, data = boundary_coord, type = "n")
  with(boundary_coord, text(x, y, 1:nrow(boundary_coord), col = "blue2", cex = 0.75))
  points(dat_loc$x, dat_loc$y, pch = 4, cex = 0.5, col = "red3")

}

# Europe with islands (points removed)
boundary_coord2 <- boundary_coord[-c(114:183, 499:503, 568:585, 587:591, 592:593),]
if (FALSE){
  plot(y~x, data = boundary_coord2, type = "n")
  with(boundary_coord2, text(x, y, 1:nrow(boundary_coord2), col = "blue2", cex = 0.75))
}

# Europe without islands (points along main coastline kept)
boundary_coord3a <- boundary_coord[c(197:544),]
boundary_coord3b <- data.frame(x = c(3761, 4803, 5320), y = c(2300, 2320, 1600))
boundary_coord3c <- boundary_coord[c(197),]
boundary_coord3 <- bind_rows(
  boundary_coord3a,
  boundary_coord3b,
  boundary_coord3c)

if (FALSE){
  plot(y~x, data = boundary_coord3, type = "n")
  with(boundary_coord3, text(x, y, 1:nrow(boundary_coord3), col = "blue2", cex = 0.75))
}

plot(y~x, data = boundary_coord3, type = "l") 
mtext("Map with ETRS89 in km, single simplified polygon")
points(dat_loc$x, dat_loc$y, pch = 4, cex = 0.5, col = "red3")

# saveRDS(boundary_coord3, "Data/104_boundary_coord3.rds")

```

### Include Ireland and "land bridge"       

```{r}

# Function for "moving" coordinates (given by column names x,y bu default)
move <- function(data, dx, dy, x="x", y="y"){
  result <- data
  result[,x] <- result[,x] + dx
  result[,y] <- result[,y] + dy
  result
}
# Test
# boundary_coord[61:63, ]
# boundary_coord[61:63, ] %>% move(100, -100)

# "Extend" Ireland
boundary_coord_ireland <- bind_rows(
  boundary_coord[52, ] %>% move(30,-20),
  boundary_coord[53:54, ],
  boundary_coord[54, ] %>% move(-20, 60), 
  # (skopping no 55)
  boundary_coord[56:58, ] %>% move(-20, 50),
  boundary_coord[59:60, ] ,
  boundary_coord[61:63, ]  %>% move(30,-20)
)

if (FALSE){
  # test plot
  plot(y~x, data = boundary_coord_ireland, col = "purple") 
  with(boundary_coord, text(x, y, 1:nrow(boundary_coord), col = "blue2", cex = 0.75))
  points(dat_loc$x, dat_loc$y, pch = 4, cex = 0.5, col = "red3")
  lines(y~x, data = boundary_coord_ireland, col = "purple") 
}

#
# Make "bridge to mainland"  
#
if (FALSE){
  plot(y~x, data = boundary_coord3, type = "n")
  with(boundary_coord3, text(x, y, 1:nrow(boundary_coord3), col = "blue2", cex = 0.75))
  with(boundary_coord_ireland, text(x, y, 1:nrow(boundary_coord_ireland), col = "blue2", cex = 0.75))
  points(dat_loc$x, dat_loc$y, pch = 4, cex = 0.5, col = "red3")

}

boundary_coord4 <- bind_rows(
  boundary_coord3[1:219,],
  boundary_coord_ireland[1:11,],
  boundary_coord3[220:352,]
)

plot(y~x, data = boundary_coord4, type = "l", col = "grey40")
with(boundary_coord4, text(x, y, 1:nrow(boundary_coord4), col = "blue2", cex = 0.75))


```


### Include Danish islands     

```{r}

if (FALSE){
  plot(y~x, data = boundary_coord4, type = "n", xlim = c(2200,3000), ylim = c(2200, 3200))
  with(boundary_coord4, text(x, y, 1:nrow(boundary_coord4), col = "blue2", cex = 0.75))
  points(dat_loc$x, dat_loc$y, pch = 4, cex = 0.5, col = "red3")
}

boundary_coord5 <- bind_rows(
  boundary_coord4[1:259,],
  boundary_coord4[324,] %>% move(-5,0),
  boundary_coord4[323,] %>% move(-10, 0),
  boundary_coord4[268:270,] %>% move(20,30),
  boundary_coord4[271:320,],
  data.frame(x = c(2876, 2834), y = c(2620, 2620)),
  boundary_coord4[322:nrow(boundary_coord4),]
)

if (F){
  plot(y~x, data = boundary_coord4, type = "n", xlim = c(2200,3000), ylim = c(2200, 3200))
  with(boundary_coord4, text(x, y, 1:nrow(boundary_coord4), col = "blue2", cex = 0.75))
  points(dat_loc$x, dat_loc$y, pch = 4, cex = 0.5, col = "red3")
  lines(boundary_coord5$x, boundary_coord5$y, col = "purple")
}

# saveRDS(boundary_coord5, "Data/104_boundary_coord5.rds")
  

plot(y~x, data = boundary_coord5, type = "n")
points(dat_loc$x, dat_loc$y, pch = 4, cex = 0.5, col = "red3")
lines(boundary_coord5$x, boundary_coord5$y, col = "purple")

```

### Remove Black Sea, include Mediterranean islands   
```{r}

if (FALSE){
  # plot(y~x, data = boundary_coord5, type = "n", xlim = c(800,3200), ylim = c(700, 1900))
  # plot(y~x, data = boundary_coord5, type = "n", xlim = c(3000, 5000), ylim = c(700, 1900))
  plot(y~x, data = boundary_coord5, type = "n", xlim = c(2500, 5500), ylim = c(700, 2400))
  # plot(y~x, data = boundary_coord5, type = "n")
  with(boundary_coord5, text(x, y, 1:nrow(boundary_coord5), col = "blue2", cex = 0.75))
  points(dat_loc$x, dat_loc$y, pch = 4, cex = 0.5, col = "red3")
}


boundary_coord6 <- bind_rows(
  boundary_coord5[1,],
  data.frame(x = c(4709, 3965, 3000, 2404, 2072, 1783), 
             y = c( 710,  609,  500,  872,  979,  884)),
  boundary_coord5[179:nrow(boundary_coord5),]
)

boundary_coord6 <- bind_rows(
  boundary_coord5[1,],
  data.frame(x = c(4709, 3965, 3000, 2404, 2072, 1783), 
             y = c( 710,  609,  500,  872,  979,  884)),
  boundary_coord5[179:354,],
  data.frame(x = c(3500, 4000, 4000, 5320),
             y = c(2050, 2050, 1700, 1700)),
  boundary_coord5[357:358,],
)

plot(y~x, data = boundary_coord6, type = "n")
# plot(y~x, data = boundary_coord6, type = "n",  xlim = c(1500, 4200) + 1000, 
#     ylim = c(400, 1900))
points(dat_loc$x, dat_loc$y, pch = 4, cex = 0.5, col = "red3")
lines(boundary_coord6$x, boundary_coord6$y, col = "purple")

# saveRDS(boundary_coord6, "Data/104_boundary_coord6.rds")

```


### Final map  
```{r}

plot(y~x, data = boundary_coord6, type = "n")
points(dat_loc$x, dat_loc$y, pch = 4, cex = 0.5, col = "red3")
lines(boundary_coord6$x, boundary_coord6$y, col = "purple")

```


## 5. Test GAM for one year, selected area

### Select data

```{r}

# 
# Scandinavia + the Baltics  
#
country_text <- "NO-SE-FI-EE-LV-LT"
countries <- strsplit(country_text, "-")[[1]]
dat2 <- dat %>% filter(countryCode %in% countries)  

dat3 <- dat2 %>% filter(year %in% 2020)  

# plot(y~x, data = boundary_coord6, type = "l")  
# points(y~x, dat3)

ggplot(dat3, aes(x,y)) +
  geom_point(aes(color = log_no3)) +
  scale_color_viridis() +
  annotate("path", x = boundary_coord6$x, y = boundary_coord6$y) +
  coord_fixed()  +
  theme_bw()


```

### Boundary

```{r}

# Make fsb (list of lists)
fsb <- list(list(v = boundary_coord6$x, w = boundary_coord6$y))

## plot boundary only
# plot(fsb[[1]]$v, fsb[[1]]$w, type="l")
# mtext("Boundary")

```

### Remove outsiders from data

```{r}

v <- dat3$x
w <- dat3$y
y <- dat3$log_no3
  
# For removing outsiders  
ind <- inSide(fsb, x=v, y=w) 
cat(round(100*mean(ind), 1), "% of the data are inside the polygon \n")

# remove outsiders
y <- y[ind]; v <- v[ind]; w <- w[ind] 
n <- length(y)

dat3_for_gam <- data.frame(v=v, w=w, y=y)

```

### Create knots

```{r}

#
# Create knots following Gavin  
#
N <- 10
gx <- seq(min(v), max(v), len = N)
gy <- seq(min(w), max(w), len = N) + 0.1  # add 0.1 in order to have fewer points on the border
# set.seed(1)
# gx <- runif(N)*5-1; gy <- runif(N)*2-1
gp <- expand.grid(gx, gy)
names(gp) <- c("x","y")

if (FALSE){
  # Check our "raw" knots
  ggplot(gp, aes(x, y)) +
    geom_point() +
    ggtitle("Knots, raw")
}

# Use inSide() to weed out knots  

# This works - x and y must be vectors with the names used
v <- gp$x
w <- gp$y
sel <- inSide(fsb, x=v, y=w) 

#if (FALSE){
  plot(fsb[[1]]$v, fsb[[1]]$w, type = "l", xlim = range(v), ylim = range(w))
  points(v, w, col = "grey50")
  text(v[sel], w[sel], 1:sum(sel), col = "blue2")
  mtext("Knots, before manual weeding")
#}

```

### Adjust knots  

- 'weed_knots' performs the 'rinse and repeat' analysis done manually above  

```{r}  
#| error: false
#| warning: false

# NOTE: Code as above, including knots adjustment  

# Make data frame for knots and start testing the GAM
# - manual weeding must take place, rinse and repeat until the GAM works  
knots_orig <- data.frame(v = v[sel], w = w[sel])

# debugonce(weed_knots)
knots <- weed_knots(knots_orig, dat3_for_gam)

plot(fsb[[1]]$v, fsb[[1]]$w, type = "l", xlim = range(v), ylim = range(w))
points(knots$v, knots$w, col = "grey50")
# text(knots$v, knots$w, 1:nrow(knots), col = "blue2")
mtext("Knots used in analysis")

```

```{r}

# GAM - dimension of the PDE solution grid
nmax <- 100 

# GAM with soap smoother
# Note that the names 'v' and 'w' must be the same for 3 objects: dat_test2, fsb and knots  
mod <- gam(y ~ s(v, w, k=30, bs="so", 
               xt = list(bnd = fsb, nmax=nmax)),
         data = dat3_for_gam,
         knots=knots)

plot(mod) ## default plot

```

## 6. Test GAM for one year, all areas

### Select data

```{r}

dat2 <- dat

dat3 <- dat2 %>% filter(year %in% 2020)  

# plot(y~x, data = boundary_coord6, type = "l")  
# points(y~x, dat3)

ggplot(dat3, aes(x,y)) +
  geom_point(aes(color = log_no3)) +
  scale_color_viridis() +
  annotate("path", x = boundary_coord6$x, y = boundary_coord6$y) +
  coord_fixed()  +
  theme_bw()


```

### Boundary

```{r}

# Make fsb (list of lists)
fsb <- list(list(v = boundary_coord6$x, w = boundary_coord6$y))

## plot boundary only
# plot(fsb[[1]]$v, fsb[[1]]$w, type="l")
# mtext("Boundary")

```

### Remove outsiders from data

```{r}

v <- dat3$x
w <- dat3$y
y <- dat3$log_no3
  
# For removing outsiders  
ind <- inSide(fsb, x=v, y=w) 
cat(round(100*mean(ind), 1), "% of the data are inside the polygon \n")

# remove outsiders
y <- y[ind]; v <- v[ind]; w <- w[ind] 
n <- length(y)

dat3_for_gam <- data.frame(v=v, w=w, y=y)

ggplot(dat3_for_gam, aes(v,w)) +
  geom_point(aes(color = y)) +
  scale_color_viridis() +
  annotate("path", x = boundary_coord6$x, y = boundary_coord6$y) +
  coord_fixed()  +
  theme_bw()

```

### Create knots

```{r}

#
# Create knots following Gavin  
#
N <- 15
gx <- seq(min(v), max(v), len = N)
gy <- seq(min(w), max(w), len = N) + 0.1  # add 0.1 in order to have fewer points on the border
# set.seed(1)
# gx <- runif(N)*5-1; gy <- runif(N)*2-1
gp <- expand.grid(gx, gy)
names(gp) <- c("x","y")

if (FALSE){
# Check our "raw" knots
ggplot(gp, aes(x, y)) +
  geom_point() +
  ggtitle("Knots, raw")
}

# Use inSide() to weed out knots  

# This works - x and y must be vectors with the names used
v <- gp$x
w <- gp$y
sel <- inSide(fsb, x=v, y=w) 

#if (FALSE){
  plot(fsb[[1]]$v, fsb[[1]]$w, type = "l", xlim = range(v), ylim = range(w))
  points(v, w, col = "grey50")
  text(v[sel], w[sel], 1:sum(sel), col = "blue2")
  mtext("Knots, before manual weeding")
#}

```

### OLD: Adjust knots and run GAM

- NOTE: Only included for possibly pedagogic reasons, now done using weed_knots()      
- Rinse and repeat analysis in order to remove knots that stil are outside land\
- As it needs to be run again and again, set k = 30 in the GAM until the knots are rinsed

```{r}

if (FALSE){
  
  # Make data frame for knots and start testing the GAM
  # - manual weeding must take place, rinse and repeat until the GAM works  
  knots <- data.frame(v = v[sel], w = w[sel])
  
  # After error in the gam: "knot 6 is on or outside boundary", add:
  knots <- knots[-6,]
  # Run again, you will get a new error: "knot 11 is on or outside boundary", add:
  knots <- knots[-11,]
  # etc.
  knots <- knots[-12,]
  knots <- knots[-33,]
  knots <- knots[-55,]
  knots <- knots[-62,]
  knots <- knots[-64,]
  knots <- knots[-66,]
  knots <- knots[-72,]
  knots <- knots[-77,]
  knots <- knots[-81,]
  
  # knots <- knots[-,]
  
  
  plot(fsb[[1]]$v, fsb[[1]]$w, type = "l", xlim = range(v), ylim = range(w))
  points(knots$v, knots$w, col = "grey50")
  text(knots$v, knots$w, 1:nrow(knots), col = "blue2")
  mtext("Knots used in analysis")
  
  
  # GAM - dimension of the PDE solution grid
  nmax <- 100 
  
  # GAM with soap smoother
  # Note that the names 'v' and 'w' must be the same for 3 objects: dat_test2, fsb and knots  
  response <- try(gam(y ~ s(v, w, k=200, bs="so", 
                            xt = list(bnd = fsb, nmax=nmax)),
                      data = dat3_for_gam,
                      knots=knots))
  
  # NOTE: should be able to make a function out of this by extracting knot nummber from "response"
  #   using some regex code
  
}

```

### Adjust knots  

- 'weed_knots' performs the 'rinse and repeat' analysis done manually above  

```{r}  
#| error: false
#| warning: false

# NOTE: Code as above, including knots adjustment  

# Make data frame for knots and start testing the GAM
# - manual weeding must take place, rinse and repeat until the GAM works  
knots_orig <- data.frame(v = v[sel], w = w[sel])

# debugonce(weed_knots)
knots <- weed_knots(knots_orig, dat3_for_gam)

plot(fsb[[1]]$v, fsb[[1]]$w, type = "l", xlim = range(v), ylim = range(w))
points(knots$v, knots$w, col = "grey50")
# text(knots$v, knots$w, 1:nrow(knots), col = "blue2")
mtext("Knots used in analysis")

```

### Adjust knots  

- 'weed_knots' performs the 'rinse and repeat' analysis done manually above  

```{r}  
#| error: false
#| warning: false

# NOTE: Code as above, including knots adjustment  

# Make data frame for knots and start testing the GAM
# - manual weeding must take place, rinse and repeat until the GAM works  
knots_orig <- data.frame(v = v[sel], w = w[sel])

# debugonce(weed_knots)
knots <- weed_knots(knots_orig, dat3_for_gam)

plot(fsb[[1]]$v, fsb[[1]]$w, type = "l", xlim = range(v), ylim = range(w))
points(knots$v, knots$w, col = "grey50")
# text(knots$v, knots$w, 1:nrow(knots), col = "blue2")
mtext("Knots used in analysis")

```

### Show model result    

```{r}

mod <- gam(y ~ s(v, w, k=200, bs="so", 
               xt = list(bnd = fsb, nmax=nmax)),
         data = dat3_for_gam,
         knots=knots)

plot(mod)

```

## 7. GAM for all years, all areas

### Select data

```{r}

dat2 <- dat

dat3 <- dat2  

# plot(y~x, data = boundary_coord6, type = "l")  
# points(y~x, dat3)

ggplot(dat3, aes(x,y)) +
  geom_point(aes(color = log_no3)) +
  scale_color_viridis() +
  annotate("path", x = boundary_coord6$x, y = boundary_coord6$y) +
  coord_fixed()  +
  theme_bw()


```

### Boundary

```{r}

# NOTE: Code as above  

# Make fsb (list of lists)
fsb <- list(list(v = boundary_coord6$x, w = boundary_coord6$y))

## plot boundary only
# plot(fsb[[1]]$v, fsb[[1]]$w, type="l")
# mtext("Boundary")

```

### Remove outsiders from data

```{r}

# NOTE: Code as above  

v <- dat3$x
w <- dat3$y
y <- dat3$log_no3
  
# For removing outsiders  
ind <- inSide(fsb, x=v, y=w) 
cat(round(100*mean(ind), 1), "% of the data are inside the polygon \n")

# remove outsiders
y <- y[ind]; v <- v[ind]; w <- w[ind] 
n <- length(y)

# Add year as wll  
dat3_for_gam <- data.frame(v=v, w=w, y=y, year = dat3$year[ind])

ggplot(dat3_for_gam, aes(v,w)) +
  geom_point(aes(color = y)) +
  scale_color_viridis() +
  annotate("path", x = boundary_coord6$x, y = boundary_coord6$y) +
  coord_fixed()  +
  theme_bw()

```

### Create knots

```{r}

# NOTE: Code as above  

#
# Create knots following Gavin  
#
N <- 15
gx <- seq(min(v), max(v), len = N)
gy <- seq(min(w), max(w), len = N) + 0.1  # add 0.1 in order to have fewer points on the border
# set.seed(1)
# gx <- runif(N)*5-1; gy <- runif(N)*2-1
gp <- expand.grid(gx, gy)
names(gp) <- c("x","y")

if (FALSE){
# Check our "raw" knots
ggplot(gp, aes(x, y)) +
  geom_point() +
  ggtitle("Knots, raw")
}

# Use inSide() to weed out knots  

# This works - x and y must be vectors with the names used
v <- gp$x
w <- gp$y
sel <- inSide(fsb, x=v, y=w) 

plot(fsb[[1]]$v, fsb[[1]]$w, type = "l", xlim = range(v), ylim = range(w))
points(v, w, col = "grey50")
text(v[sel], w[sel], 1:sum(sel), col = "blue2")
mtext("Knots, before manual weeding")


```


### Function for adjusting knots   

- Performs the rinse and repeat procedure done above  

```{r}

weed_knots <- function(knots, data){
  
  # 'knots' is expected to be a dataframe with variables v and w  
  # 'data' is expected to be a dataframe with variables v, w and y  
  # 'fsb' is expected to exist (a list of lists, containing border info)   
  
  old.options <- options(show.error.messages = FALSE, warn = -1)

  knot_error <- TRUE
  
  while (knot_error){
    response <- try(
      # The following gam is hard-coded - adjust depending on what you need
      # The last time the gam will run (when the knots have been weeded so it actually works),
      #   so a quite low k is suggested to save time  
      gam(y ~ s(v, w, k=30, bs="so", 
                xt = list(bnd = fsb, nmax=100)),
          data = data,
          knots=knots)
    )
    if (class(response) != "try-error"){
      knot_error <- FALSE
    } else {
      knotno_char <- stringr::str_extract(response[1], "(?<=knot\\s)[0-9]+")
      if (!is.na(knotno_char)){
        knotno <- as.numeric(knotno_char)
        knots <- knots[-knotno,]
      } else {
        knot_error <- FALSE  # it's an error, but not knot_error....
      }
    }
  }
  
  options(old.options)
  
  knots

}

```

### Adjust knots  

- 'weed_knots' performs the 'rinse and repeat' analysis done manually above  

```{r}  
#| error: false
#| warning: false

# NOTE: Code as above, including knots adjustment  

# Make data frame for knots and start testing the GAM
# - manual weeding must take place, rinse and repeat until the GAM works  
knots_orig <- data.frame(v = v[sel], w = w[sel])

# debugonce(weed_knots)
knots <- weed_knots(knots_orig, dat3_for_gam)

plot(fsb[[1]]$v, fsb[[1]]$w, type = "l", xlim = range(v), ylim = range(w))
points(knots$v, knots$w, col = "grey50")
# text(knots$v, knots$w, 1:nrow(knots), col = "blue2")
mtext("Knots used in analysis")

```


### Run model      

```{r}

mod <- gam(y ~ s(v, w, k=200, bs="so", 
               xt = list(bnd = fsb, nmax=nmax)),
         data = dat3_for_gam,
         knots=knots)

# Error in soap.basis(sd, x, y, film = TRUE, wiggly = TRUE, penalty = TRUE) : \n  NA/NaN/Inf in foreign function call (arg 13)\n  

plot(mod)

```

### MAnual adjustment also needed  
```{r}

plot(fsb[[1]]$v, fsb[[1]]$w, type = "l", xlim = range(v), ylim = range(w))
points(knots$v, knots$w, col = "grey50")
text(knots$v, knots$w, 1:nrow(knots), col = "blue2")
mtext("Knots used in analysis")

knots <- knots[-11,]

mod <- gam(y ~ s(v, w, k=200, bs="so", 
               xt = list(bnd = fsb, nmax=nmax)),
         data = dat3_for_gam,
         knots=knots)

plot(mod)

```


### With year effect

```{r}

if (FALSE){
  tic()
  nmax = 500
  mod2 <- gam(y ~ s(v, w, k=500, bs="so", 
                    xt = list(bnd = fsb, nmax=nmax)) + s(year, k = 5),
              data = dat3_for_gam,
              knots=knots)
  toc()   # 89 sec with nmax = 500 and k = 500  
  
  saveRDS(mod2, "Data/104_mod2.rds")
}

mod2 <- readRDS("Data/104_mod2.rds")

plot(mod2) ## default plot


```


```{r}

dat3_loc <- dat3_for_gam %>%
  

```

### Predikert  
```{r}

df_pred <- data.frame(v = 2000, w = 2000, year = 1992:2020)  
pred <- predict(mod2, df_pred, se.fit = TRUE)

df_pred$log_no3 <- pred$fit

df_pred$no3 <- 10^pred$fit
df_pred$no3_lo <- 10^(pred$fit - pred$se.fit)
df_pred$no3_up <- 10^(pred$fit + pred$se.fit)

ggplot(df_pred, aes(year, no3)) +
  geom_ribbon(aes(ymin = no3_lo, ymax = no3_up), fill = "lightblue") +
  geom_line()

```


### Year effect only

```{r}

mod3a <- gam(y ~ s(year, k = 5),
              data = dat3_for_gam)
plot(mod3a) 

```
```{r}

visreg::visreg(mod3a)

```

### Year effect only

```{r}

mod3b <- gam(y ~ s(v, w, k=100) + s(year, k = 5),
              data = dat3_for_gam)
plot(mod3b) 


```